\vspace{1.0cm} \Large \bf Supplemental material of ``Health and disease imprinted in the time variability of the human microbiome''

### 2016-04-17 mSystems/mSys_MandM.tex
MINOR: rename mySys_ file prefix to mSys_ prefix
On branch master
	renamed:    mySys_MAIN.pdf -> mSys_MAIN.pdf
	renamed:    mySys_MAIN.tex -> mSys_MAIN.tex
	renamed:    mySys_MandM.tex -> mSys_MandM.tex
	renamed:    mySys_discussion.tex -> mSys_discussion.tex
	renamed:    mySys_intro.tex -> mSys_intro.tex
	renamed:    mySys_results.tex -> mSys_results.tex
	renamed:    mySys_suppl_0.1.tex -> mSys_suppl_0.1.tex

\documentclass[11pt,oneside,letterpaper]{article}

% graphicx package, useful for including eps and pdf graphics
\usepackage{graphicx}
\DeclareGraphicsExtensions{.pdf,.png,.jpg}

% basic packages
\usepackage{color} 
\usepackage{parskip}
\usepackage{float}

% text layout
\usepackage{geometry}
\geometry{textwidth=15cm} % 15.25cm for single-space, 16.25cm for double-space
\geometry{textheight=22cm} % 22cm for single-space, 22.5cm for double-space

% helps to keep figures from being orphaned on a page by themselves
\renewcommand{\topfraction}{0.85}
\renewcommand{\textfraction}{0.1}
%% Roman footnote numbering style to not collide with bibliographic cites: 
\renewcommand{\thefootnote}{\Roman{footnote}}

\newcommand{\CC}[0]{\emph{cmplxcruncher}}
\newcommand{\task}[1]{\texttt{\bfseries\scshape\textcolor{MidnightBlue}{#1}}}
\newcommand{\un}[1]{\operatorname{#1}}
\newcommand{\unclassrate}[0]{\un{kpb/s/core}}

% bold the 'Figure #' in the caption and separate it with a period
% Captions will be left justified
\usepackage[labelfont=bf,labelsep=period,font=small]{caption}

% review layout with double-spacing
\usepackage{setspace} 
\doublespacing
\captionsetup{labelfont=bf,labelsep=period,font=doublespacing}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}
\renewcommand\citeleft{(}
\renewcommand\citeright{)}
\renewcommand\citeform[1]{\textsl{#1}}

% Remove brackets from numbering in list of References
\renewcommand\refname{\large References}
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

\usepackage{authblk}
\renewcommand\Authands{ \& }
\renewcommand\Authfont{\normalsize \bf}
\renewcommand\Affilfont{\small \normalfont}
\makeatletter
\renewcommand\AB@affilsepx{, \protect\Affilfont}
\makeatother

% notation
\usepackage{amsmath}
\usepackage{amssymb}

%%% TITLE %%%
\title{\vspace{1.0cm} \Large \bf 
Supplemental material of ``Health and disease imprinted in the time variability of the human microbiome''
}

\author[1]{Jose Manuel Mart\' i}
\author[1,2]{Daniel M. Mart\' inez}
\author[1]{Manuel Pe\~na}
\author[1]{C\'esar Gracia}
\author[2,3,4]{Amparo Latorre}
\author[2,3,4]{Andr\'es Moya}
\author[1]{Carlos P. Garay}

\affil[1]{Instituto de Fisica Corpuscular, CSIC-UVEG, P.O.  22085, 46071, Valencia, Spain.}
\affil[2]{FISABIO, Avda de Catalunya, 21, 46020, Valencia, Spain.}
\affil[3]{Cavanilles Institute of Biodiversity and Evolutionary Biology, Univ. de Valencia, 46980, Spain. }
\affil[4]{CIBER en Epidemiolog\' ia y Salud P\'ublica (CIBEResp), Madrid, Spain}

\date{}

\begin{document}

\maketitle

\section{Model} \label{sec:model}

We model the microbial abundances across time along the lines of Blumm \textit{et al.}\cite{ranking}. The dynamics of taxon relative abundances is described by the Langevin equation: 
\begin{equation}
\dot{x_i} = F_i \cdot x_i^\alpha + V \cdot x_i^\beta \xi_i(t) - \phi(t) \cdot x_i,
\end{equation}
where F$_i$ captures the fitness of the taxon i, V corresponds to the noise amplitude and $\xi_i$(t) is a Gaussian random noise with zero mean  $<\xi_i(t)>$ = 0 and variance uncorrelated in time, $<\xi_i(t) \xi_i(t')>$ =  $\delta(t'-t)$, . The function $\phi(t)$ ensures the normalization at all times, $\sum x_i(t) = 1$, and corresponds to $\phi(t) = \sum F_i x_i^\alpha + \sum V x_i^\beta \xi_i(t)$ .
The temporal evolution of the probability that a taxon i has a relative abundance $x_i(t)$, P(x$_i$,t), is determined by the Fokker-Planck equation:
\begin{equation}
\frac{\partial P}{\partial t} = - \frac{\partial}{\partial x_i}  [(F_i \cdot x_i^\alpha - \phi(t) \cdot x_i ) \cdot P]+ \frac{1}{2} \frac{\partial^2}{\partial x_i^2} (V^2 \cdot x_i^{2\beta}\cdot P).
\end{equation}
The microbiota evolves towards a steady-state with a time-independent probability depending on the values of $\alpha$, $\beta$, F$_i$ and V. For $\alpha<1$ (otherwise, systems are always unstable), the steady-state probability may be localized in a region around a preferred value or broadly distributed over a wide range, depending on whether the fitness F$_i$  dominates or is overwhelmed by the noise amplitude V. The steady-state solution of the Fokker-Planck equation is given by:
\begin{eqnarray}
P_0 (x_i) &=& C_{ne}(\alpha,\beta,F_i,V)  \cdot x_i^{-2\beta}  \cdot \exp\Big[\frac{2F_i}{V^2}\frac{x_i^{1+\alpha-2\beta}}{1+\alpha-2\beta}-\frac{\phi_0}{V^2}\frac{x_i^{2-2\beta}}{1-\beta}\Big] \quad \textrm{if} \quad  2\beta \ne 1+\alpha, \notag \\
P_0 (x_i) &=& C_e(\alpha,\beta,F_i,V)  \cdot x_i^{\frac{2F_i}{V^2} -2\beta}  \cdot \exp\Big[\frac{\phi_0}{V^2}\frac{x_i^{2-2\beta}}{1-\beta}\Big] \quad \textrm{if} \quad  2\beta = 1+\alpha,
\notag
\end{eqnarray}
where $\phi_0 = (\sum_i F_i^{1/(1-\alpha)})^{1-\alpha}$ and C$_{ne}$ and C$_{e}$ are integrals that should be solved numerically for the parameters of interest. The ordered phase happens when the solution has a maximum in the physical interval ($0<x_i<1$). For larger V, the transition to a disordered phase happens when the maximum shifts to the unphysical region $x_i<0$, which sets the phase transition region V($\alpha,\beta,F_i$). The phase transition region can be calculated analytically in particular cases:
\begin{eqnarray}
F_i^2 &=& 4 \beta \phi_0 V^2\quad \textrm{if} \quad  \beta = \alpha \neq 1, \notag \\
F_i &=& \beta V^2\quad \textrm{if} \quad  2\beta = 1+\alpha,
\notag
\end{eqnarray}
where the first case, simplifies to $F= 3 V^2$ if $\beta = 0.75$ and the fitness of this taxon dominates in $\phi_0$. 
In many physical systems (Brownian motion is the classical example), the two terms of the Langevin equation are related.  The \emph{fluctuation--dissipation theorem} states a general relationship between the response to an external disturbance and the internal fluctuations of the system\cite{FD}. The theorem can be used as the basic formula to derive the fitness from the analysis of fluctuations 
of the microbiota, assuming that it is in equilibrium (the ordered phase).

\section{Selection and Methods}

\subsection{Sample selection}
We have chosen studies about relevant pathologies containing metagenomic sequencing time data series of bacterial populations from humans in different healthy and non-healthy states. We have selected only those individuals who had three or more time points of data available in databases. Metadata of each study is provided in Supplementary Tables \ref{tab:diet} to \ref{tab:LEA}. All used 16S rRNA gene sequencing except for the study of the discordant kwashiorkor twins\cite{kwashiorkor} (see Supplementary Tables \ref{tab:DH} and \ref{tab:DK}) where shotgun metagenomic sequencing (SMS) and 16S rRNA were used. In the latter case we selected to work with SMS data to show that our method is valid regardless of the source of taxonomic information. Each one of the datasets was treated as follows:

\subsection{16rRNA sequences processing}
Reads from the selected studies were first quality filtered using the FastX toolkit\cite{FASTX}, allowing only those reads which had more than 25 of quality along the 75\% of the complete sequence. 16S rRNA reads were then clustered at 97\% nucleotide sequence identity (97\% ID) into operational taxonomic units (OTUs) using QIIME package software\cite{QIIME} (version 1.8) We followed open reference OTU picking workflow in all cases. The clustering method used was uclust, and the OTUs were matched against Silva database\cite{SILVA} (version 111, July 2012) and were assigned to taxonomy with an uclust-based consensus taxonomy assigner. The parameters used in this step were: similarity 0.97, prefilter percent id 0.6, max accepts 20, max rejects 500.

\subsection{Metagenomic sequences processing}
Metagenomic shotgun (and 16S too) sequences were analyzed with LMAT (Livermore Metagenomics Analysis Toolkit) software package\cite{LMAT} (version 1.2.4, with Feb'15 release of data base \emph{LMAT-Grand}). LMAT was run using a Bull shared-memory node belonging to the team's HPC (high performance computing) cluster. It is equipped with 32 cores (64 threads available using Intel Hyper-threading technology) as it has 2 Haswell-based Xeons, the E5-2698v3@2.3 GHz, sharing half a tebibyte (0.5 TiB, that is, 512 gibibytes) of DRAM memory. This node is also provided with a card PCIe SSD as NVRAM, the P420m HHHL, with 1.4 TB, and 750000 reading IOPS, 4 KB, achieving 3.3 GB/s, which Micron kindly issued free of charge, as a sample for testing purposes. The computing node was supplied with a RAID-0 (striping) scratch disk area. We used the ``Grand'' database\footnote{In this context, ``Grand'' refers to a huge database that contains k-mers from all viral, prokaryote, fungal and protist genomes present in the NCBI database, plus Human reference genome (hg19), plus GenBank Human, plus the 1000 Human Genomes Project (HGP). This represent about 31.75 billion k-mers occupying 457.62 GB.}, release Feb'15, provided by the LMAT team. Previously to any calculation, the full database was loaded in the NVRAM. With this configuration the observed LMAT sustained sequence classification rate was $20\unclassrate$. Finally, it is worth mentioning that a complete set of Python scripts have been developed as back-end and front-end of the LMAT pipeline in order to manage the added complexity of time series analysis.

\subsection{Taxa level selection}
We selected genus as taxonomic level for the subsequent steps of our work. In order to ensure that, between adjacent taxonomic levels, there were not crucial differences which could still be of relevance after standardization (see Section \ref{sec:stan}), we tested two different data sets. In the former, the antibiotics study\cite{antibiotic} with 16S data, we tested the differences between genus and family levels. The latter dataset tested was the kwashiorkor discordant twins study\cite{kwashiorkor} for both genus and species taxonomic levels. The Supplementary Figures \ref{fig:taxlev1} (overview) and \ref{fig:taxlev2} (detail) plot the comparison between studies (and so, 16S and SMS) and between adjacent taxonomic levels.

@@

\section{Temporal evolution of model parameters}

We have studied the time dependence of the variability $V$ and power law index $\beta$ (see Section \ref{sec:model}) by using a sliding window approach. The total number of time points are divided in subsets of five points, where next subset is defined by adding next time sampling and by eliminating the earliest one. Both parameters were calculated for each subset against the average time lapse. Figure \ref{fig:tempevo1} shows the variability  $V$ as a function of time for the largest sampling: two individuals in the Caporaso's study\cite{moving} corresponding to the gut microbiota of a male (upper plot) and a female (lower plot). Figure \ref{fig:tempevo2} shows the time evolution of $V$ for patient P2 of the IBS study\cite{IBS} (upper plot) and patient D in the antibiotics study\cite{antibiotic} (lower plot).

\begin{figure}
	\includegraphics[width=1.0\textwidth]{results/sliwin/male_mov.pdf}
	\hspace*{3mm}\includegraphics[width=0.448\textwidth]{results/sliwin/female_mov.pdf}
\caption{$V$ as a function of time for the two individuals in the Caporaso's study\cite{moving}: samples of gut microbiome of a male (upper plot) and a female (lower plot). Both samples show changes in the variability V with quasi--periodic behavior peaked at about 10 days. Variability grows more for the gut microbiota of the male and share a minimal value around 0.1 with the gut microbiota of the female.}
\label{fig:tempevo1}
\end{figure}

\begin{figure}
	\centering 
 	\includegraphics[width=0.8\textwidth]{results/sliwin/patP2_IBS.pdf}
  	\includegraphics[width=1.0\textwidth]{results/sliwin/patD_antibio.pdf} 
\caption{$V$ as a function of time for patient P2 of the IBS study\cite{IBS} (upper plot) and patient D in the antibiotics study\cite{antibiotic} (lower plot). The variability 
of the gut microbiota of P2 decreases from above 0.3 to below 0.2, showing a slow tendency to increase the order of the system.  Antibiotic intake leaks to a quick increase of variability which lasts for a few days to recover ordering. The second antibiotic treatment shows some memory (lower increase of variability) with a slower recovery. NOTE: The blue vertical lines in the lower plot are showing the periods of antibiotic treatment.}
\label{fig:tempevo2}
\end{figure}

\section{Engineered Software}

A complete software framework, named 'ComplexCruncher', has been engineered to support the analysis of the dynamics of ranking processes in complex systems. Although the software was devised with a clear bias towards metagenomics, it is general enough to be able to cope with a ranking process in any complex system. Implemented in Python using well-known open-source community software, the software solution is composed of two parts that can be used together or apart: a web-based graphic front-end connected to a database, and a computing kernel. Used together, this software enables other users to reproduce our results easily and, furthermore, upload and analyse their own data or experiment with the preloaded metagenomics data sets. The next sections of this supplementary material deal with these two software pieces in detail.

\section{Web-based front-end}

@@

CCWebPortal is a web application that runs on current versions of many browsers. Additional software is not needed and only requires javaScript to be enabled on the browser to run applications. CCWebPortal is, at least, compatible with the following browsers: Internet Explorer 6+, Firefox 3.6+ (PC, Mac), Safari 4+, Chrome 10+ and Opera 11+ (PC, Mac).

CCWebPortal is implemented following the client$-$server distributed programming model. The user connects to the website and a set of javaScript files are downloaded and executed on the local machine. Subsequently, the client application connects to a remote server that enables the execution of calculations and transactions through a centralized database management system. The web application is deployed on the client side using ExtJs libraries. ExtJs is a javaScript development framework to implement interactive Web applications using AJAX, DHTML and DOM technologies. ExtJS has been widely adopted by large technology projects worldwide. Its success lies in its ability to design and implement web interfaces with a clear and flexible functionality for the end user. ExtJs is used under the GNU license GPL v3.

The CCWebPortal website communicates with a remote server that hosts a MySQL database. The database runs on a MyISAM engine that optimizes the speed of access to records in a transactional web-based environment. A set of relational tables allows the structuring of the metagenomics repository to establish relationships between records. Thus the search and information threshing is optimized for queries launched from the client interface. Access to the database on the server is implemented through Django framework. Django is an open-source framework written in Python using the model-view-controller (MVC) architectural pattern for implementing user interfaces, which eases the development and updating of code. This framework is used to implement web applications on the server side and allows flexible and modular developments and access to various types of database managers. Django runs as a web server, awaiting requests from the client CCWebPortal application. Once a user is authorized to perform a given operation, Django interacts with the database serving queries to the client application.

\section{Computing kernel}

The effective data analysis has been performed with a Python tool developed from scratch following the OOP\footnote{Object Oriented Programming, but OOP also stands for Object Oriented Python, which is applicable here too.} paradigm, called \CC. This software is the back-end of the website described in the previous subsection. However, it could be run as an independent piece of software since it is built as a Python package provided with a command-line front-end (\CC.py). Once installed, the tool can be run interactively (using the flag \texttt{-i}/\texttt{--interactive}) but also in automatic mode (flag \texttt{-a}/\texttt{--automatic} followed by the desired graphic format, e.g., \texttt{pdf}, \texttt{png}, \texttt{svg}, \texttt{ps}), which uses parallel computation to speed up the analysis of several data sources.

The external mandatory dependencies of \CC\ are with Python packages \emph{pandas}, \emph{matplotlib}, \emph{numpy}, \emph{scipy} and \emph{xlrd}, while the recommended ones are \emph{pylatex} and \emph{uncertainties}. The PEP8 compliant code of \CC\ could be executed under a Python 2.7 interpret, but a 3.X interpret is strongly advised. The complete distribution of \CC\ has the code signature shown in the Table \ref{tab:cloc}, obtained through the \emph{CLOC}\footnote{\emph{Count Lines of Code} is a perl script that counts blank lines, comment lines, and physical lines of source code.} tool release 1.62. Figure \ref{fig:CC} shows the code layout of the complete distribution of \CC.

\begin{table}
  \begin{center}
    \begin{tabular}{lrrrrcrcr}
    \hline
	Language & files & blank & comment & code & x & scale & = & 3rd gen. equiv \\
	\hline
	Python & 13 & 384 & 666 & 4222 &x& 4.20 &=& 17732.40 \\
	R & 1 & 11 & 38 & 192 &x& 3.00 &=& 576.00 \\	
	\hline
	SUM: & 14 & 395 & 704 & 4414 &x& 4.15 &=& 18308.40 \\
	\hline
    \end{tabular}
  \end{center}
  \caption{Line counting for the complete distribution of \CC\ by \emph{CLOC}}
  \label{tab:cloc}
\end{table}

\begin{figure}
  \centering
	\[ \text{\task{ComplexCruncher}}
	\begin{cases}
		\text{\task{cmplxcruncher.py}} \\
		\\ 
		\text{\task{cmplxcruncher}}
		\begin{cases}
			\text{\task{\_\_init\_\_.py}} \\ 
			\\ 
			\text{\task{models.py}} \\
			\\ 
			\text{\task{plots.py}} \\
			\\ 
			\text{\task{sessions.py}} \\ 
		\end{cases}
		\\ \\
		\text{\task{tools}}
		\begin{cases}
			\text{\task{power\_analysis.R}} \\
			\\ 
			\text{\task{rawlmat2lmat.py}} \\
			\\ 
			\text{\task{lmat2cmplx.py}} \\
			\\ 
			\text{\task{pyLMAT\_rl.py}} \\
			\\ 
			\text{\task{pyLMAT\_cs.py}} \\
			\\ 
			\text{\task{pyLMAT\_gl.py}} \\
			\\
			\text{\task{LMAT}}
			\begin{cases}
	    		\text{\task{pyLMAT\_rescore.py}} \\ 
	    		\\ 
	    		\text{\task{HMPspiker.py}} 
			\end{cases}
		\end{cases}
	\end{cases} \]
	\caption{\CC\ layout}
	\label{fig:CC}
\end{figure}

@@

\section{Un-weighted power-law fit}\label{sec:unw}

\subsection{Fitting the best model}

As mentioned above, to choose between fitting power laws ($y=Vx^\beta$) using linear regression on log-transformed (LLR) data versus non-linear regression (NLR), we mainly follow \emph{General Guidelines for the Analysis of Biological Power Laws}\cite{ecology}. It consists of the following three steps:

@@

@@

@@

\subsection{Calculating the coefficient of determination}

>>>>

We have analysed more than 35000 time series of taxa from the gut microbiomes of 97 individuals (sampling from three up to 332 time points), obtained from publicly available high throughput sequencing data on: healthy individuals over a long term span\cite{moving}, people with various degrees of obesity\cite{LEA}, twin pairs discordant for kwashiorkor\cite{kwashiorkor}, response to diet changes\cite{diet} or to antibiotic perturbation\cite{antibiotic}, and subjects diagnosed with irritable bowel syndrome (IBS)\cite{IBS}. 
We engineered a complete software framework and a web platform, ComplexCruncher, ready to be implemented by other users.

We summarize our dataset in ST1-6 (Tables 1-6 in suplemental material). The bacteria and archaea taxonomic assignations were obtained by analysing 16S rRNA sequences, which were clustered into operational taxonomic units (OTUs) sharing 97 \% sequence identity using QIIME\cite{QIIME}. WGS data\cite{kwashiorkor} were analysed and assigned at strain level by the Livermore Metagenomic Analysis Toolkit (LMAT)\cite{LMAT}, according to their default quality threshold. Genus, with best balance between error assignment and number of taxa, was chosen as our reference taxonomic level. We have verified that our conclusions are not significantly affected by selecting family or species as the reference taxonomic level (see Figure 1 in supplemental material).

\todo{Specify, in each study treated, the nature of the samples (conditions, timespan between timepoints, subjects). Specify, and it is very important, what we consider ?healthy? in each study (for example: pre-antibiotics is healthy)}

%%%
\subsection*{Model} \label{sec:model}

We model the microbial abundances across time along the lines of Blumm \textit{et al.}\cite{ranking}. The dynamics of taxon relative abundances is described by the Langevin equation:
\begin{linenomath}
\begin{equation}
\dot{x_i} = F_i \cdot x_i^\alpha + V \cdot x_i^\beta \xi_i(t) - \phi(t) \cdot x_i,
\end{equation}
\end{linenomath}
where F$_i$ captures the fitness of the taxon i, V corresponds to the noise amplitude and $\xi_i$(t) is a Gaussian random noise with zero mean  $<\xi_i(t)>$ = 0 and variance uncorrelated in time, $<\xi_i(t) \xi_i(t')>$ =  $\delta(t'-t)$, . The function $\phi(t)$ ensures the normalization at all times, $\sum x_i(t) = 1$, and corresponds to $\phi(t) = \sum F_i x_i^\alpha + \sum V x_i^\beta \xi_i(t)$.
The temporal evolution of the probability that a taxon i has a relative abundance $x_i(t)$, P(x$_i$,t), is determined by the Fokker-Planck equation:
\begin{linenomath}
\begin{equation}
\frac{\partial P}{\partial t} = - \frac{\partial}{\partial x_i}  [(F_i \cdot x_i^\alpha - \phi(t) \cdot x_i ) \cdot P]+ \frac{1}{2} \frac{\partial^2}{\partial x_i^2} (V^2 \cdot x_i^{2\beta}\cdot P).
\end{equation}
\end{linenomath}
The microbiota evolves towards a steady-state with a time-independent probability depending on the values of $\alpha$, $\beta$, F$_i$ and V. For $\alpha<1$ (otherwise, systems are always unstable), the steady-state probability may be localized in a region around a preferred value or broadly distributed over a wide range, depending on whether the fitness F$_i$  dominates or is overwhelmed by the noise amplitude V. The steady-state solution of the Fokker-Planck equation is given by:
\begin{eqnarray}
P_0 (x_i) &=& C_{ne}(\alpha,\beta,F_i,V)  \cdot x_i^{-2\beta}  \cdot \exp\Big[\frac{2F_i}{V^2}\frac{x_i^{1+\alpha-2\beta}}{1+\alpha-2\beta}-\frac{\phi_0}{V^2}\frac{x_i^{2-2\beta}}{1-\beta}\Big] \quad \textrm{if} \quad  2\beta \ne 1+\alpha, \notag \\
P_0 (x_i) &=& C_e(\alpha,\beta,F_i,V)  \cdot x_i^{\frac{2F_i}{V^2} -2\beta}  \cdot \exp\Big[\frac{\phi_0}{V^2}\frac{x_i^{2-2\beta}}{1-\beta}\Big] \quad \textrm{if} \quad  2\beta = 1+\alpha,
\notag
\end{eqnarray}
where $\phi_0 = (\sum_i F_i^{1/(1-\alpha)})^{1-\alpha}$ and C$_{ne}$ and C$_{e}$ are integrals that should be solved numerically for the parameters of interest. The ordered phase happens when the solution has a maximum in the physical interval ($0<x_i<1$). For larger V, the transition to a disordered phase happens when the maximum shifts to the unphysical region $x_i<0$, which sets the phase transition region V($\alpha,\beta,F_i$). The phase transition region can be calculated analytically in particular cases:
\begin{eqnarray}
F_i^2 &=& 4 \beta \phi_0 V^2\quad \textrm{if} \quad  \beta = \alpha \neq 1, \notag \\
F_i &=& \beta V^2\quad \textrm{if} \quad  2\beta = 1+\alpha,
\notag
\end{eqnarray}
where the first case, simplifies to $F= 3 V^2$ if $\beta = 0.75$ and the fitness of this taxon dominates in $\phi_0$. 
In many physical systems (Brownian motion is the classical example), the two terms of the Langevin equation are related.  The \emph{fluctuation--dissipation theorem} states a general relationship between the response to an external disturbance and the internal fluctuations of the system\cite{FD}. The theorem can be used as the basic formula to derive the fitness from the analysis of fluctuations 
of the microbiota, assuming that it is in equilibrium (the ordered phase).

\todo{Explain better the fluctiation-dissipation theorem}

%%%
\subsection*{Selection and Methods}

\subsubsection*{Sample selection}
We have chosen studies about relevant pathologies containing metagenomic sequencing time data series of bacterial populations from humans in different healthy and non-healthy states. We have selected only those individuals who had three or more time points of data available in databases. Metadata of each study is provided in Supplementary Tables \ref{tab:diet} to \ref{tab:LEA}. All used 16S rRNA gene sequencing except for the study of the discordant kwashiorkor twins\cite{kwashiorkor} (see Supplementary Tables \ref{tab:DH} and \ref{tab:DK}) where shotgun metagenomic sequencing (SMS) and 16S rRNA were used. In the latter case we selected to work with SMS data to show that our method is valid regardless of the source of taxonomic information. Each one of the datasets was treated as follows:

\subsubsection*{16rRNA sequences processing}
Reads from the selected studies were first quality filtered using the FastX toolkit\cite{FASTX}, allowing only those reads which had more than 25 of quality along the 75\% of the complete sequence. 16S rRNA reads were then clustered at 97\% nucleotide sequence identity (97\% ID) into operational taxonomic units (OTUs) using QIIME package software\cite{QIIME} (version 1.8) We followed open reference OTU picking workflow in all cases. The clustering method used was uclust, and the OTUs were matched against Silva database\cite{SILVA} (version 111, July 2012) and were assigned to taxonomy with an uclust-based consensus taxonomy assigner. The parameters used in this step were: similarity 0.97, prefilter percent id 0.6, max accepts 20, max rejects 500.

\subsubsection*{Metagenomic sequences processing}
Metagenomic shotgun (and 16S too) sequences were analyzed with LMAT (Livermore Metagenomics Analysis Toolkit) software package\cite{LMAT} (version 1.2.4, with Feb'15 release of data base \emph{LMAT-Grand}). LMAT was run using a Bull shared-memory node belonging to the team's HPC (high performance computing) cluster. It is equipped with 32 cores (64 threads available using Intel Hyper-threading technology) as it has 2 Haswell-based Xeons, the E5-2698v3@2.3 GHz, sharing half a tebibyte (0.5 TiB, that is, 512 gibibytes) of DRAM memory. This node is also provided with a card PCIe SSD as NVRAM, the P420m HHHL, with 1.4 TB, and 750000 reading IOPS, 4 KB, achieving 3.3 GB/s, which Micron kindly issued free of charge, as a sample for testing purposes. The computing node was supplied with a RAID-0 (striping) scratch disk area. We used the ``Grand'' database\footnote{In this context, ``Grand'' refers to a huge database that contains k-mers from all viral, prokaryote, fungal and protist genomes present in the NCBI database, plus Human reference genome (hg19), plus GenBank Human, plus the 1000 Human Genomes Project (HGP). This represent about 31.75 billion k-mers occupying 457.62 GB.}, release Feb'15, provided by the LMAT team. Previously to any calculation, the full database was loaded in the NVRAM. With this configuration the observed LMAT sustained sequence classification rate was $20\unclassrate$. Finally, it is worth mentioning that a complete set of Python scripts have been developed as back-end and front-end of the LMAT pipeline in order to manage the added complexity of time series analysis.

\subsubsection*{Taxa level selection}
We selected genus as taxonomic level for the subsequent steps of our work. In order to ensure that, between adjacent taxonomic levels, there were not crucial differences which could still be of relevance after standardization (see Section \ref{sec:stan}), we tested two different data sets. In the former, the antibiotics study\cite{antibiotic} with 16S data, we tested the differences between genus and family levels. The latter dataset tested was the kwashiorkor discordant twins study\cite{kwashiorkor} for both genus and species taxonomic levels. The Supplementary Figures \ref{fig:taxlev1} (overview) and \ref{fig:taxlev2} (detail) plot the comparison between studies (and so, 16S and SMS) and between adjacent taxonomic levels.

@@

\subsection*{ComplexCruncher}

A complete software framework, named 'ComplexCruncher', has been engineered to support the analysis of the dynamics of ranking processes in complex systems. Although the software was devised with a clear bias towards metagenomics, it is general enough to be able to cope with a ranking process in any complex system. Implemented in Python using well-known open-source community software, the software solution is composed of two parts that can be used together or apart: a web-based graphic front-end connected to a database, and a computing kernel. Used together, this software enables other users to reproduce our results easily and, furthermore, upload and analyse their own data or experiment with the preloaded metagenomics data sets.

@@

CCWebPortal is a web application that runs on current versions of many browsers. Additional software is not needed and only requires javaScript to be enabled on the browser to run applications. 
CCWebPortal is implemented following the client$-$server distributed programming model, where the javaScript client application connects to a remote server that enables the execution of calculations and transactions through a centralized database management system. A set of relational tables allows the structuring of the metagenomics repository to establish relationships between records. Thus the search and information threshing is optimized for queries launched from the client interface. Access to the database on the server is implemented through Django framework, an open-source framework written in Python using the model-view-controller (MVC) architectural pattern for implementing user interfaces.

The effective data analysis has been performed with a Python tool developed from scratch to more than 4200 lines of code. Implemented following the Object Oriented Programming, paradigm, this software is the back-end of the website described above. However, it could be run as an independent piece of software since it is built as a Python package provided with a command-line front-end (\CC.py). Once installed, the tool can be run interactively but also in automatic mode, which uses parallel computation to speed up the analysis of several data sources.

@@

\subsection*{Un-weighted power-law fit}\label{sec:unw}

\subsubsection*{Fitting the best model}

As already mentioned, to choose between fitting power laws ($y=Vx^\beta$) using linear regression on log-transformed (LLR) data versus non-linear regression (NLR), we mainly follow \emph{General Guidelines for the Analysis of Biological Power Laws}\cite{ecology}. It consists of the following three steps:

@@

            \begin{linenomath}

@@

            \end{linenomath}

@@

\subsubsection*{Calculating the coefficient of determination}

We think the best approach in this situation is to apply the generalized $R^2$ that, for continuous models, was defined as\cite{genR2}:
\begin{linenomath}
$$ R^2 = 1 - \left(\frac{\mathcal{L}(0)}{\mathcal{L}(\hat\theta)}\right)^{\!\!\frac{2}{n}} $$
\end{linenomath}
where $\mathcal{L}(\hat\theta)$ and $\mathcal{L}(0)$  denote the likelihoods of the fitted and the ``null'' model, respectively, and $n$ is the sample size. In terms of the loglikelihoods, the generalized coefficient of determination would be:
\begin{linenomath}
$$ R^2 = 1 - \mathrm{e}^{-\frac{2}{n}\left(\log\mathcal{L}(\hat\theta)-\log\mathcal{L}(0)\right)} $$
\end{linenomath}
We have the likelihoods calculated from the previous section, but what about the ``null'' models? We understand that they are the models with only the intercept. So for the Gaussian additive error model:
\begin{linenomath}
$$\mathcal{L}_\text{norm}(0) = \prod_{i=1}^n\left[\frac{1}{\sqrt{2\pi\sigma^2_\text{NLR0}}}\;\exp{\left(-\frac{\left(y_i-\bar y\right)^2}{2\sigma^2_\text{NLR0}}\right)}\right]$$
\end{linenomath}
So:
\begin{eqnarray*}
\log\mathcal{L}_\text{norm}(0) &=& -\frac{n}{2}\log\left|2\pi\sigma^2_\text{NLR0}\right| - \frac{1}{2\sigma^2_\text{NLR0}}\sum_{i=1}^n\left(y_i-\bar y\right)^2\\
 &=& -\frac{n}{2}\left(\log\left|2\pi\sigma^2_\text{NLR0}\right| + 1 \right)
\end{eqnarray*}
since $\sigma^2_\text{NLR0}=\frac{1}{n}\sum\left(y_i-\bar y\right)^2=\frac{1}{n}\mathrm{TSS}_\text{NLR}$. Now, coming back to the coefficient of determination, we have:
\begin{eqnarray*}
R^2_\text{NLR} &=& 1 - \mathrm{e}^{\frac{2}{n}\left(\log\mathcal{L}_\text{NLR}(0)-\log\mathcal{L}_\text{NLR}(\hat\theta)\right)} = 1 - \exp{\left(\frac{\log(\mathrm{RSS}_\text{NLR})}{\log(\mathrm{TSS}_\text{NLR})}\right)}
 =\\ &=& 1 - \frac{\mathrm{RSS}_\text{NLR}}{\mathrm{TSS}_\text{NLR}} =
 1 - \frac{\sum_{i=1}^n\left(y_i-V_\text{NLR}x_i^{\beta_\text{NLR}}\right)^2}{\sum_{i=1}^n\left(y_i-\bar y\right)^2}
\end{eqnarray*}
recovering the traditional expression for $R^2$. Using the same approach for calculating $R^2_\text{LLR}$, then:
\begin{linenomath}
$$\mathcal{L}_\text{logn}(0) = \prod_{i=1}^n\left[\frac{1}{y_i\sqrt{2\pi\sigma^2_\text{LLR0}}}\;\exp{\left(-\frac{\left(\log|y_i|-\log|B_\text{LLR0}|\right)^2}{2\sigma^2_\text{LLR0}}\right)}\right]$$
\end{linenomath}
So:
\begin{eqnarray*}
\log\mathcal{L}_\text{logn}(0) &=& -\frac{n}{2}\log\left|2\pi\sigma^2_\text{LLR0}\right| - \frac{1}{2\sigma^2_\text{LLR0}}\sum_{i=1}^n\left(\log|y_i|-\overline{\log|y|}\right)^2 - \sum_{i=1}^n\log|y_i|\\
 &=& -\frac{n}{2}\left(\log\left|2\pi\sigma^2_\text{LLR0}\right| + 1 \right) - \sum_{i=1}^n\log|y_i|
\end{eqnarray*}
since $\sigma^2_\text{LLR0}=\frac{1}{n}\sum\left(\log|y_i|-\overline{\log|y|}\right)^2=\frac{1}{n}\mathrm{TSS}_\text{logn}$. Again, recalling the expression for the generalized coefficient of determination, we have:
\begin{eqnarray*}
R^2_\text{LLR} &=& 1 - \mathrm{e}^{\frac{2}{n}\left(\log\mathcal{L}_\text{LLR}(0)-\log\mathcal{L}_\text{LLR}(\hat\theta)\right)} = 1 - \exp{\left(\frac{\log(\mathrm{RSS}_\text{LLR})}{\log(\mathrm{TSS}_\text{LLR})}\right)}
 =\\ &=& 1 - \frac{\mathrm{RSS}_\text{LLR}}{\mathrm{TSS}_\text{LLR}} =
 1 - \frac{\sum_{i=1}^n\left(\log|y_i|-\log|V_\text{LLR}|-\beta_\text{LLR}\log|x_i|\right)^2}{\sum_{i=1}^n\left(\log|y_i|-\overline{\log|y|}\right)^2}
\end{eqnarray*}

\subsection*{X-weighted power-law fit}\label{sec:X-w}
